---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

# ducksemantics

`ducksemantics` is a DuckDB-native semantic and retrieval substrate for R. It
owns ontology graphs, validated HPO observations, cataloged Monarch relations,
snapshot-bound literature retrieval, source-grounded judgments, and optional
local model/provider protocols. It does not own ClinVar/PubMed source storage,
case ranking, or evaluation.

Bulk values are ordinary data frames or caller-owned DuckDB relations. S7 is
reserved for real prompt, embedding, parser, and annotator provider protocols.
See [ARCHITECTURE.md](ARCHITECTURE.md) for the current boundary.

## Install

```r
install.packages(
  "ducksemantics",
  repos = c("https://rgenomicsetl.r-universe.dev", "https://cloud.r-project.org")
)
```

## Ontology candidates and accepted HPO observations

Lexical candidates are intentionally distinct from accepted observations. An
accepted observation must carry its HPO identifier, exact zero-based half-open
source span, context, method/provider provenance, confidence, and explicit
`accepted` status.

```{r hpo}
library(ducksemantics)

documents <- data.frame(
  document_id = "case-001",
  source_text = "The patient has seizures.",
  stringsAsFactors = FALSE
)

observations <- data.frame(
  document_id = "case-001",
  hpo_id = "HP:0001250",
  start_offset = 16L,
  end_offset = 24L,
  source_text = "seizures",
  context_status = "present",
  method = "lexical_alias",
  provider_id = "local-lexicon",
  provider_version = "1",
  confidence = 1,
  status = "accepted",
  stringsAsFactors = FALSE
)

ducksemantics_hpo_observations(documents, observations)
```

The validator rejects hallucinated text, empty spans, out-of-bounds spans, and
any context outside `present`, `absent/negated`, `family_history`, `uncertain`,
`conflict`, or `unsupported`.

## Cataloged Monarch facts

Callers provide both facts and a typed release catalog. A catalog row identifies
one provider/release and supplies a `Date`/`POSIXct` effective date or a numeric
source ordinal. Exact and as-of queries bind the provider plus a cataloged
release; historical gene-disease holdouts are explicit anti-join audits.

```{r monarch}
releases <- data.frame(
  release_id = c("2024-01", "2025-01"),
  provider_id = "Monarch",
  effective_date = as.Date(c("2024-01-01", "2025-01-01")),
  source_ordinal = c(1, 2),
  stringsAsFactors = FALSE
)
gene_disease <- data.frame(
  gene_id = c("HGNC:1100", "HGNC:1100"),
  disease_id = c("MONDO:0000001", "MONDO:0000002"),
  release_id = c("2024-01", "2025-01"),
  provider_id = "Monarch",
  stringsAsFactors = FALSE
)

gene_disease <- ducksemantics_monarch_import(
  gene_disease, releases, relation = "gene_disease"
)
ducksemantics_monarch_gene_disease_holdout_audit(
  gene_disease, releases, provider_id = "Monarch",
  train_release_id = "2024-01", holdout_release_id = "2025-01"
)
```

## Attached dated Monarch annotation packs

For an official dated Monarch DuckDB pack already available to DuckDB, attach it
outside the package, verify its immutable receipt, and bind one exact typed
catalog row. The projection does
not attach URLs, download or collect pack rows into R, make a copy, or accept
`latest`; it creates seven connection-local `TEMP VIEW`s over the attached pack.
The role views retain every raw edge and add normalized roles only when explicit
Biolink subject/object categories support the orientation. The source contract
matches Monarch's serialized text `negated` field and also projects
`node_has_phenotype`. Filter on a `supported_*` `association_status`
deliberately: malformed categories or negation, missing predicates, negated
edges, and unsupported orientations are returned as statuses, never silently
treated as support or causal evidence. `negation_status` still distinguishes an
omitted optional qualifier from explicit false, explicit true, and malformed
text; omission does not erase Monarch's positive source assertion.

```{r monarch-pack, eval=FALSE}
conn <- ducksemantics_connect()
DBI::dbExecute(
  conn,
  "ATTACH '/data/monarch-kg/2026-07-14/monarch-kg.duckdb' AS monarch (READ_ONLY)"
)

releases <- data.frame(
  provider_id = "infores:monarchinitiative",
  release_id = "2026-07-14",
  effective_date = as.Date("2026-07-14"),
  stringsAsFactors = FALSE
)
views <- ducksemantics_monarch_project_pack(
  conn, source_catalog = "monarch", releases = releases,
  provider_id = "infores:monarchinitiative", release_id = "2026-07-14"
)

DBI::dbGetQuery(
  conn,
  "SELECT gene_id, disease_id, predicate, primary_knowledge_source
   FROM semantic_monarch_gene_disease
   WHERE association_status IN (
     'supported_subject_to_object', 'supported_object_to_subject'
   )"
)
```

The returned `views` relation names views rather than materializing pack
contents. The native association view preserves edge identifiers, predicate and
category, primary/aggregator provenance, evidence/publication/qualifier arrays,
agent and knowledge fields, taxa, negation, and raw original fields. No
provider edges are voted or merged.

## Snapshot-bound literature retrieval

`RClinVarbitration` owns append-only PubMed source history and
`ducksemantics` neither imports nor stores literature. Retrieval consumes the
same provider-scoped source-order projection:

- **snapshots:** `provider_id`, `snapshot_id`, finite integer
  `high_water_ordinal`, and optional `POSIXct` `effective_at`;
- **article versions:** `provider_id`, `article_id`, `pmid`, `version_id`,
  finite integer `source_ordinal`, and logical `is_deleted`;
- **version-bound sections:** `provider_id`, `article_id`, `pmid`,
  `version_id`, finite integer `source_ordinal`, `section`, and exact non-empty
  `text`.

For a cataloged provider/snapshot, it selects the maximum
`source_ordinal <= high_water_ordinal` per article, excludes an article when
that selected event is deleted, and joins sections only on all version keys.
Thus order is numeric (for example 10 follows 9), not lexical. Exact nonempty
zero-based half-open source spans are returned.

This is directly derivable from RClinVarbitration without a storage adapter:
`pubmed_sources` supplies provider, snapshot, source ordinal, and application
timestamp; `pubmed_articles` supplies the event and deletion flag; and
`pubmed_abstracts` supplies same-source version-bound sections. A caller may
also project `article_title` into a `title` section. The relations remain
caller-owned; `ducksemantics` does not write or materialize a shadow copy.

```{r literature}
snapshots <- data.frame(
  provider_id = c("pubmed", "pubmed"),
  snapshot_id = c("baseline", "update"),
  high_water_ordinal = c(9, 10),
  effective_at = as.POSIXct(c("2025-01-09", "2025-01-10"), tz = "UTC"),
  stringsAsFactors = FALSE
)
articles <- data.frame(
  provider_id = "pubmed",
  article_id = c("pmid:123", "pmid:123"),
  pmid = "123",
  version_id = c("source-9", "source-10"),
  source_ordinal = c(9, 10),
  is_deleted = c(FALSE, FALSE),
  stringsAsFactors = FALSE
)
sections <- data.frame(
  provider_id = "pubmed",
  article_id = c("pmid:123", "pmid:123"),
  pmid = "123",
  version_id = c("source-9", "source-10"),
  source_ordinal = c(9, 10),
  section = "BACKGROUND",
  text = c("Seizure phenotype in a family", "Updated seizure phenotype"),
  stringsAsFactors = FALSE
)

ducksemantics_literature_retrieve(
  articles, sections, snapshots,
  query = "seizure",
  provider_id = "pubmed",
  snapshot_id = "update"
)
```
